
\abstract*{Acknowledgements}

\extrachap{Acknowledgements}

Mark and Ang Davis extended an invitation to give a talk in the pub for Bollington Science Festival in 2012 which started me wondering why it wasn't possible to explain why plants had Fibonacci numbers, even though Keith Devlin had written twenty years earlier that this was a mathematically solved problem. Around the same time, Erinma Ochu, then of the MOSI Turing's Sunflowers project, was amazed not only that no-one knew how to breed Lucas sunflowers, but that I was surprised by the question.  
I'm particularly grateful to Julia Gog who made some usefully mathematical remarks on a very early draft of the lattice theory chapters, and to Paul Glendinning who long ago gave some advice on how to phrase my challenge to the Fundamental Theorem of Phyllotaxis.
Christopher Golé generously gave me some snatched time and told me about his recent work with Stéphane Douady on the stability of rhombic lattices with three disks. Stéphane, together with  Annemiek Cornelissen made me welcome on a visit to the CNRS Laboratoire Matière et Systèmes Complexes. Tamsin Spelman and several of her colleagues at the Sainsbury Laboratory made helpful comments, and Hugo Tavares and Katie Abley's neighbourly support was particularly nourishing. Adam Swinton has been a patient and critical reader of several drafts.   Philip Maini graciously hosted me in the Mathematical Biology seminar at the University of Oxford. and Andrew Krause, Mark Muldoon and Ronjoy Adhikari listened patiently and constructively to my pitches for this material. Phil Ramsden currently holds the record for the fastest finding of a mistake.  

Some years ago,  Diana Gillooly and David Tranah offered supportive, engaged and productive publication advice over a rather extended period. It was only after I had given up trying to interest a publisher in this book that Alain Goriely on his own initiative invited me to join the Springer Texts series, and I thank him and the Editorial Board for support for this book. I thank Christopher Golé, Stéphane Douady and Richard Schwartz for their constructive advice as reviewers for Springer. 
%I expect to be able to write that Donna Chernyk shepherded this through the press at Springer Nature with friendly efficiency.


This book has been many years in the writing, and love of mathematics alone could never have got it finished. For the love of almost everything else, including getting things done, I thank in particular Emma Anderson. But it was the questions of \textit{all} these people, by email, in the pub, the museum and the seminar room, that shaped this attempt at answers.  Thanks to you all. 


