
\thispagestyle{titlingpage}

\chapter*{To the reader}
This draft is incomplete and needs help from novice and expert alike:
\begin{itemize}
	\item If you like the theorem-proof style of mathematics then Chapters 3-5 need checking for logical consistency, and the examples need testing (including whether they are too easy/hard/irrelevant). I'd like to hear about the use of pictures too. If you don't, then do the summaries at the end of these chapters explain what you need?
	\item By contrast, the second half covers a much broader range of material, is much less formal, and currently less coherent. Here I want to know what you think about the choice of topics, and how intelligible their coverage is to a non-expert. 
	\item If you're already an expert, my rule has been not to cite papers whose point I don't understand. I don't intend to publish this book until my Zotero folder of papers I don't know if I understand or not is empty, but that is currently bulging. Please be patient (or better, explanatory).
\item \textbf{Please don't pass on to anyone else.} Be warned that copyright information on the images is labile, unreliable, and incomplete. 
 \copynote\ means an image is probably in copyright and copyright permissions will need to be secured. 
 \placenote\ means an image which will later be replaced by another one. 
 \item Happy to hear about typos/design, but don't sweat over them. 
 \item  You can email comments to \href{mailto:jonathan@swintons.net}{jonathan@swintons.net}. Please mention that this is draft \jdraftnumber.
 \end{itemize}

\noindent%
Thanks,\\
Jonathan Swinton\\
Manchester, January 2021\\
%\listoftodos
\clearpage



